Issues in developing multivariable models for treatment selection

### 2016-05-10 Documents/paper.tex
lisas edits

---
title: "Issues in developing multivariable models for treatment selection"
author: "Michael C Sachs and Lisa M McShane"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
    pdf_document:
      citation_package: natbib
      keep_tex: yes
      fig_caption: yes
      pandoc_args: ["--bibliography", "paper.bib"]
fontsize: 12pt
geometry: margin=1in
---

# Abstract

@@

# Introduction

```{r setup, include = FALSE}
library(dplyr)
library(ggplot2)
library(knitr)
library(survival)
library(ggkm)
library(tidyr)

opts_chunk$set(echo = FALSE, message = FALSE, error = FALSE, warning = FALSE)
knit_hooks$set(plot = function(x, options) {
  if ('tikz' %in% options$dev && !options$external) {
    hook_plot_tex(x, options)
  } else hook_plot_md(x, options)
})
```

Omics technologies that generate a large amount of molecular data about biospecimens have the potential to provide information about a patient's disease characteristics above and beyond standard clinical and pathological features. By combining the information from a large amount of molecular features into a multivariable model, hereafter referred to as a _biomarker signature_, there is the opportunity to identify distinct subgroups of patients for whom treatment decisions can be personalized. A multivariable biomarker can guide the decisions to treat or not to treat and help identify the patients who are most likely to survive. The key challenge we address in this paper is how to combine features from a high dimensional molecular assay to derive a signature that is fit for a specified clinical use and to provide valid estimates of performance characteristics.

## Terminology and Notation

A **biomarker signature** is a transformation of multiple individual features, typically molecular characterstics measured on a multiplex assay, to a one-dimensional space. Specifically, let $X$ denote the set of $p$ features under consideration. The signature is an unknown function $f(X): \mathbb{R}^p \mapsto \mathbb{R}^1$. The signature result may be continuous, take multiple discrete values, or be dichotomous.

Let $S$ denote the development dataset which includes, for each of $n$ represented individuals, a feature vector $X$, an outcome $Y$, a treatment $Z$, and possibly other variables. $S$ is a sample of size $n$ from distribution $\mathcal{P}$ with domain $\mathcal{S}$. Let $\mathcal{F}$ be a mapping from $\mathcal{X}$ to the space of continuous functions with domain $\mathbb{R}^p$ and range $\mathbb{R}$, $\mathcal{D}$. Thus $\mathcal{F}: \mathcal{X} \mapsto \mathcal{D}$ denotes the process or algorithm through which a particular $f$ is estimated. We do not place any other restrictions on $\mathcal{F}$, it could be a clustering approach, a regression approach, a combination of both, or something else entirely. We will use $\mathcal{F}$ to denote the manner in which $f$ is estimated and will write $f \in \mathcal{F}$ to denote that $f$ is estimated with the class of methods $\mathcal{F}$.

Let $\phi: \mathcal{D} \times \mathcal{X} \mapsto \mathbb{R}$ denote the statistic that quantifies the performance of the function $f$, such as predictive accuracy, mean squared error, or area under the receiver operating characteristic (ROC) curve (AUC). This could also be a measure of association, such as an odds ratio, hazard ratio, or log-rank statistic. This is a function of both $f$ and $S$. We are interested in estimating $E_\mathcal{P}[\phi_{f^*}(S)]$, which is the expected error under the data generation mechanism, for a particular $f^* \in \mathcal{F}$. This allows us to understand how the signature will perform on future observations generated from $\mathcal{P}$. We may also be interested in estimating $E_\mathcal{P}[\phi_f(S)]$ *for all* $f \in \mathcal{F}$, which is the generalization error for $f$ generated using mechanism $\mathcal{F}$. This doesn't guide outside researchers as to which specific $f$ to use, yet it is useful for development because it tells us how much signal is in the data. As shorthand we will write this as $E_\mathcal{P}[\phi_\mathcal{F}(S)]$

A signature that **reliably predicts** an outcome $Y$ is one that has generalization error small enough for the clinical context. Such a signature may be useful for treatment selection, prognosis, or other type of clinical management.

## Overview of biomarker signature development

The main goal is to produce a good signature. To establish this requires a valid estimate of the signature's performance. Assessment of the worth of a biomarker signature requires providing a valid estimate of the performance of $f \in \mathcal{F}$. Additionally, one can provide a specification of $f$ for others to use. Typically, a specific $f$ is estimated using $\mathcal{F}$ based on some training data. This can be done using a variety of different methods. In recent years there have been an explosion in the literature of computational approaches to classification and prediction, and we do not intend to summarize them all here. Some excellent reviews of such approaches are @hastie2009elements and @moons2012riskI. The main considerations in signature estimation are identifying the features to include, deciding what transformations to apply, determining how to combine the feature measurements, and whether to apply thresholds/cutoffs to the resulting signature value.

@@

Signatures can be estimated by identifying naturally occuring clusters, or intrinsic subtypes using $X$ alone without regard to the outcome $Y$. The PAM-50 gene signature was developed using clustering methods in an attempt to identify intrinsic subtypes with different prognosis [@parker2009supervised]. Later, the intrinsic subtypes were shown to be strongly associated with clinical outcomes [@nielsen2010comparison]. Supervised learning techniques can also be used to identify prognostic signatures, as was the case with Oncotype DX [@paik2004multigene], another signature developed for clinical decision making in early stage breast cancer. In this case, regression-based methods are used to estimate a model that is highly predictive for the outcome $Y$. In the case of Oncotype DX, the outcome in question was distant recurrence of breast cancer.

An approach to identifying treatment-selection signatures is to use regression techniques to estimate a signature that has a strong interaction with a particular treatment. It is possible, and quite common in high-dimensional settings, to combine multiple approaches to estimating $f$. For instance, a data-reduction step by variable selection or clustering may be performed before doing regression analysis on the resulting components.

No matter what the particular model building method is, our main concern and focus of this paper is with obtaining a valid estimate of its performance, that is, a good estimate of $E_\mathcal{P}[\phi_\mathcal{F}(S)]$. This depends on the true signal in the data and the specific algorithm $\mathcal{F}$ used. An optional component of the development phase is to provide a specification of $f$ for others to use on independent data or in clinical practice.

## Biomarker signatures in clinical practice

@@

## Data analysis example

Throughout this paper, we reanalyze data from @zhu2010prognostic. Briefly, the data of interest are from the JBR.10 trial, which was a randomized controlled trial of adjuvant vinorelbine/cisplatin (ACT) versus observation alone (OBS) in 482 participants with non small cell lung cancer (NSCLC). Of those 482 participants, 169 had frozen tissue collected, and of those samples, 133 (71 in ACT and 62 in OBS) had gene-expression profiling performed using U133A oligonucleotide microarrays (Affymetrix, Santa Clara, CA).

The goal of the @zhu2010prognostic paper was to identify a multi-gene signature that strongly predicts prognosis, and the hypothesis was that the poor prognosis subgroup would benefit more from ACT compared to the good prognosis subgroup. The signature was trained to predict disease specific survival. The annotated gene expression data and clinical information are available from the Gene Expression Omnibus (identifier: GSE14814, @edgar2002gene).

@zhu2010prognostic present results that mainly focus on the discrimination ability of their estimated signature. They do that by demonstrating that the two risk subgroups predicted by their signature (high risk and low risk) have separation in their survival curves and that the hazard ratio for their signature is large and significant even when adjusting for other risk factors. They do not directly address calibration, that is, whether their signature accurately predicts survival times.

We used a similar approach to preprocessing as did @zhu2010prognostic, as we could not reproduce their workflow exactly due to outdated software. Batch effects were removed using the `ComBat` function in the `sva` `R` package [@leeksva] and then the gene expression values were centered by their means and scaled by their standard deviations. Our signature development approach is similar but not identical to that in @zhu2010prognostic. For purposes of illustrating the concepts we used a simplyfied approach to signature development that retains the main features of the original. The exact approach to signature development does not have a major impact on the main conclusions of our evaluation of the various approaches to signature performance assessment.

@@

# Issues

Recall that the main goal is to estimate $E_\mathcal{P}[\phi_\mathcal{F}(S)]$, the expected value of a given statistic on future observations for $f \in \mathcal{F}$. This can be estimated with the in-sample empirical estimate: $\hat{E}[\phi_f(S)] = \frac{1}{n}\sum_{i=1}^n\phi_f(s_i)$ for a particular $f$. However, if $S$ is used to estimate $f$ then the estimate will be biased due to overfitting, that is, $|E_\mathcal{P}[\phi_f(S)] - \hat{E}[\phi_f(S)]|$ will be large. This bias results from the fact that $\phi$ depends on $f$, and thus the statistic $\phi$ is being adaptively defined based on the observed data $S$. Overfitting occurs when a model is fit to noise in the data. This often occurs when fitting a model that is overly complex relative to the amount of signal in the available data.

In many cases during signature development, the performance metric $\phi$ in question is a statistic that relates to calibration. While a biomarker signature may accurately predict a clinical outcome, that does not neccessarily imply that the signature is clinically useful. To assess discrimination, a different statistic $\phi$ may be used, such as the area under the ROC curve. Measures of association such as the odds ratio, hazard ratio, or difference in survival probabilities are examples of performance metrics that are commonly used, although their value for biomarker signatures has been debated [@pepe2004limitations]. Evaluation of $\phi$ is subject to bias due to overfitting regardless of what $\phi$ is used, a fact that is commonly overlooked in the medical literature.

In @zhu2010prognostic, $\phi$ was the hazard ratio comparing the high risk and low risk groups as defined by the JBL.10 signature $f$. The risk groups were determined by the signature $f$, which was estimated using $S$, the same data that was then used to produce the hazard ratio estimate of 15. They go on to assess the performance of the JBL.10 signature in a series of independent data sets, in which they estimate hazard ratios of approximately 2. Here we reanalyze the dataset and illustrate some remedies to avoid bias in estimating signature performance.

## Avoiding Overfitting

A traditional method to avoid overfitting is the split sample approach. First, randomly partition $S$ into the training sample $S_t$ and the holdout sample $S_h$ with sample sizes $n_t$ and $n_h$, respectively. Then, $S_h$ is hidden from the analyst while $\mathcal{F}$ is applied to $S_t$ to estimate the signature function $f_t$. For fixed $f_t$, $\hat{E}[\phi_{f_t}(S_h)]$ is an unbiased estimator of both $E_{\mathcal{P}}[\phi_{f_t}(S)]$ and $E_\mathcal{P}[\phi_\mathcal{F}(S)]$. The specific form of $f_t$ that is fixed using the $S_t$ partition can be reported as the function for others to use, therefore the aforementioned estimator is also an estimate of the error for that specific $f_t$. The drawback of the split-sample approach is that the performance of $f_t$ is likely to be inferior to the performance of a function $f^*$ derived using the same approach applied to the entire data set. @dobbin2011optimally investigate how to optimally split a dataset into training and holdout partitions.

Another approach to avoid overfitting is cross-validation, which is a resampling based approach. For a fixed integer $k$, which can be between 1 and $n$, we randomly select a partition of $k$ observations from $S$, denoted $S_k$. Then $f_{-k}$ is estimated and fixed by appling $\mathcal{F}$ on $S_{-k}$ which is the subset of $S$ that is disjoint from $S_k$. Then, we get an estimate $\hat{E}[\phi_{f_{-k}}(S_k)]$ which is an unbiased estimate of $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$. This process is repeated $K$ times to yield $K$ estimates. Each of these estimates is unbiased, but noisy, because typically $k$ is very small relative to $n$. Thus, we average over $K$ to get a less noisy estimate. This process is called "leave $k$ out" or "$n/k$ fold" cross-validation. Note that for each partition that is selected, we obtain a new estimate of $f$, therefore we are only estimating $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$ and not the performance for a single specified signature. Typically, if a specific form for $f$ is desired, it would be estimated using the entire dataset $S$ to yield $f^*$ as above.

A variation on the cross-validation approach is bootstrapping. In that case, a sample $S_b$ of size $n$ is sampled _with replacement_ from $S$. Then $f_{b}$ is estimated and fixed by applying $\mathcal{F}$ to $S_{b}$. The performance metric $\phi$ is calculated on the subset of $S$ that is disjoint from $S_b$: $S_{-b}$ to yield an estimate of $E[\phi_{f_b}(S_{-b})]$. This process is repeated $K$ times to yield $K$ estimates. These $K$ estimates of the performance metrics are averaged to obtain the mean over the bootstrap replicates. @efron1997improvements suggest a variation, the 0.632 estimate: 
$$
\hat{E}^*[\phi_{\mathcal{F}}(S)] = .368 \hat{E}[\phi_{f}(S)] + 0.632 \hat{E}[\phi_{f_b}(S_{-b})],
$$
where $\hat{E}[\phi_{f}(S)]$ is the naive estimate of $\phi_f$ using the entire dataset.

Another variation on all of these methods is the concept of pre-validation [@tibshirani2002pre]. With pre-validation, instead of computing the statistic $\phi$ for each of the held-out subsets ($S_{-b}$ for the bootstrap or $S_{k}$ for cross-validation), the fitted predictor $\hat{f}(X_i)$ is estimated for $X_i \in S_{-b}$ where $\hat{f}$ is estimated using $S_{b}$. This process is repeated to obtain a set of pre-validated predictor estimates $\hat{f}$ which are then used to calculate $\phi$. For single-step holdout, this process is equivalent to what is described above. For cross-validation and the bootstrap, this process avoids the problem of having too few cases to estimate the statistic $\phi$ on the held-out dataset.

## Simulation Study

@@

To illustrate the different properties of these estimates and how they deal with overfitting, we conduct a simulation study. Data were generated with 1000 observations, each with a binary outcome $Y$ with prevalence 0.3, and 500 features sampled from the standard normal distribution. This is the null case where no features are associated with $Y$. The signature development procedure entails a feature selection step, in which each feature is regressed against $Y$ in a univariate logistic regression model. The 25 features with the smallest p-values are selected for inclusion in a multivariable logistic regression model which defines the final signature.

We compare each of the methods described above, split-sample holdout, cross-validation, bootstrap, and pre-validation, along with several commonly used but biased approaches. All of the approaches are described in \ref{descript}. Two of the biased approaches use the full sample to select the features, followed by fitting the multivariable model on the holdout subset. This is referred to as "parital holdout" or "partial CV" when using split-sample holdout or cross-validation as the validation step, respectively. We also implemented the naive resubstitution approach, wherein the model is trained and evaluated on the same dataset, and the partial resubstitution approach wherein the model is trained on a holdout set and then evaluated on the combined complete dataset. Our main interest is in comparing the bias and variance of the resulting estimates of $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$. In our simulation, we look at two different performance metrics, the area under the ROC curve (AUC) and the odds ratio for the outcome comparing the signature groups.

```{r cvsims, fig.cap = "Comparison of different approaches to estimating the Area Under the ROC Curve (AUC) in the setting where a dataset is used to both develop the signature and evaluate its performance. The violin plots show mirrored density estimates for the AUC for 1000 replicates of the numerical experiment. In each replicate, there are 1000 observations and 500 features. The true value of the AUC is 0.5. CV = Cross validation. \\label{fig1}"}
load("../Code/cvsim-result.RData")
cvres <- do.call("rbind", cvsims)
cvlong <- do.call("rbind", lapply(1:ncol(cvres), function(i){

  cl <- cvres[, i]
  splow <- strsplit(colnames(cvres)[i], ".", fixed = TRUE)
  nm <- paste(unlist(sapply(splow, function(s) rev(rev(s)[-1]))), collapse = ".")
  cls <- sapply(splow, function(s) rev(s)[1])
  data.frame(value = cvres[, i], stat = cls, scen = nm, stringsAsFactors = FALSE)

}))
cvlong$value[cvlong$stat == "OR"] <- exp(cvlong$value[cvlong$stat == "OR"])

cvlong$scen <- factor(cvlong$scen, levels = c("naive", "zhu.cv", "zhu.hold", "zhu.hold2", 
                                              "cv.preval", "cv.10", "cv.100", "holdout.3", 
                                              "holdout.5", "boot"), ordered = TRUE)

>>>>

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{natbib}
\bibliographystyle{plainnat}
\usepackage{longtable,booktabs}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={Michael C Sachs and Lisa M McShane},
            pdftitle={Issues in developing multivariable models for treatment selection},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}
  \title{Issues in developing multivariable models for treatment selection}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{Michael C Sachs and Lisa M McShane}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{May 10, 2016}

\begin{document}

\maketitle

\section{Abstract}\label{abstract}

@@

\section{Introduction}\label{introduction}

Omics technologies that generate a large amount of molecular data about
biospecimens have the potential to provide information about a patient's
disease characteristics above and beyond standard clinical and
pathological features. By combining the information from a large amount
of molecular features into a multivariable model, hereafter referred to
as a \emph{biomarker signature}, there is the opportunity to identify
distinct subgroups of patients for whom treatment decisions can be
personalized. A multivariable biomarker can guide the decisions to treat
or not to treat and help identify the patients who are most likely to
survive. The key challenge we address in this paper is how to combine
features from a high dimensional molecular assay to derive a signature
that is fit for a specified clinical use and to provide valid estimates
of performance characteristics.

\subsection{Terminology and Notation}\label{terminology-and-notation}

A \textbf{biomarker signature} is a transformation of multiple
individual features, typically molecular characterstics measured on a
multiplex assay, to a one-dimensional space. Specifically, let \(X\)
denote the set of \(p\) features under consideration. The signature is
an unknown function \(f(X): \mathbb{R}^p \mapsto \mathbb{R}^1\). The
signature result may be continuous, take multiple discrete values, or be
dichotomous.

Let \(S\) denote the development dataset which includes, for each of
\(n\) represented individuals, a feature vector \(X\), an outcome \(Y\),
a treatment \(Z\), and possibly other variables. \(S\) is a sample of
size \(n\) from distribution \(\mathcal{P}\) with domain
\(\mathcal{S}\). Let \(\mathcal{F}\) be a mapping from \(\mathcal{X}\)
to the space of continuous functions with domain \(\mathbb{R}^p\) and
range \(\mathbb{R}\), \(\mathcal{D}\). Thus
\(\mathcal{F}: \mathcal{X} \mapsto \mathcal{D}\) denotes the process or
algorithm through which a particular \(f\) is estimated. We do not place
any other restrictions on \(\mathcal{F}\), it could be a clustering
approach, a regression approach, a combination of both, or something
else entirely. We will use \(\mathcal{F}\) to denote the manner in which
\(f\) is estimated and will write \(f \in \mathcal{F}\) to denote that
\(f\) is estimated with the class of methods \(\mathcal{F}\).

Let \(\phi: \mathcal{D} \times \mathcal{X} \mapsto \mathbb{R}\) denote
the statistic that quantifies the performance of the function \(f\),
such as predictive accuracy, mean squared error, or area under the
receiver operating characteristic (ROC) curve (AUC). This could also be
a measure of association, such as an odds ratio, hazard ratio, or
log-rank statistic. This is a function of both \(f\) and \(S\). We are
interested in estimating \(E_\mathcal{P}[\phi_{f^*}(S)]\), which is the
expected error under the data generation mechanism, for a particular
\(f^* \in \mathcal{F}\). This allows us to understand how the signature
will perform on future observations generated from \(\mathcal{P}\). We
may also be interested in estimating \(E_\mathcal{P}[\phi_f(S)]\)
\emph{for all} \(f \in \mathcal{F}\), which is the generalization error
for \(f\) generated using mechanism \(\mathcal{F}\). This doesn't guide
outside researchers as to which specific \(f\) to use, yet it is useful
for development because it tells us how much signal is in the data. As
shorthand we will write this as \(E_\mathcal{P}[\phi_\mathcal{F}(S)]\)

A signature that \textbf{reliably predicts} an outcome \(Y\) is one that
has generalization error small enough for the clinical context. Such a
signature may be useful for treatment selection, prognosis, or other
type of clinical management.

\subsection{Overview of biomarker signature
development}\label{overview-of-biomarker-signature-development}

The main goal is to produce a good signature. To establish this requires
a valid estimate of the signature's performance. Assessment of the worth
of a biomarker signature requires providing a valid estimate of the
performance of \(f \in \mathcal{F}\). Additionally, one can provide a
specification of \(f\) for others to use. Typically, a specific \(f\) is
estimated using \(\mathcal{F}\) based on some training data. This can be
done using a variety of different methods. In recent years there have
been an explosion in the literature of computational approaches to
classification and prediction, and we do not intend to summarize them
all here. Some excellent reviews of such approaches are
\citet{hastie2009elements} and \citet{moons2012riskI}. The main
considerations in signature estimation are identifying the features to
include, deciding what transformations to apply, determining how to
combine the feature measurements, and whether to apply
thresholds/cutoffs to the resulting signature value.

@@

Signatures can be estimated by identifying naturally occuring clusters,
or intrinsic subtypes using \(X\) alone without regard to the outcome
\(Y\). The PAM-50 gene signature was developed using clustering methods
in an attempt to identify intrinsic subtypes with different prognosis
\citep{parker2009supervised}. Later, the intrinsic subtypes were shown
to be strongly associated with clinical outcomes
\citep{nielsen2010comparison}. Supervised learning techniques can also
be used to identify prognostic signatures, as was the case with Oncotype
DX \citep{paik2004multigene}, another signature developed for clinical
decision making in early stage breast cancer. In this case,
regression-based methods are used to estimate a model that is highly
predictive for the outcome \(Y\). In the case of Oncotype DX, the
outcome in question was distant recurrence of breast cancer.

An approach to identifying treatment-selection signatures is to use
regression techniques to estimate a signature that has a strong
interaction with a particular treatment. It is possible, and quite
common in high-dimensional settings, to combine multiple approaches to
estimating \(f\). For instance, a data-reduction step by variable
selection or clustering may be performed before doing regression
analysis on the resulting components.

No matter what the particular model building method is, our main concern
and focus of this paper is with obtaining a valid estimate of its
performance, that is, a good estimate of
\(E_\mathcal{P}[\phi_\mathcal{F}(S)]\). This depends on the true signal
in the data and the specific algorithm \(\mathcal{F}\) used. An optional
component of the development phase is to provide a specification of
\(f\) for others to use on independent data or in clinical practice.

\subsection{Biomarker signatures in clinical
practice}\label{biomarker-signatures-in-clinical-practice}

@@

\subsection{Data analysis example}\label{data-analysis-example}

Throughout this paper, we reanalyze data from \citet{zhu2010prognostic}.
Briefly, the data of interest are from the JBR.10 trial, which was a
randomized controlled trial of adjuvant vinorelbine/cisplatin (ACT)
versus observation alone (OBS) in 482 participants with non small cell
lung cancer (NSCLC). Of those 482 participants, 169 had frozen tissue
collected, and of those samples, 133 (71 in ACT and 62 in OBS) had
gene-expression profiling performed using U133A oligonucleotide
microarrays (Affymetrix, Santa Clara, CA).

The goal of the \citet{zhu2010prognostic} paper was to identify a
multi-gene signature that strongly predicts prognosis, and the
hypothesis was that the poor prognosis subgroup would benefit more from
ACT compared to the good prognosis subgroup. The signature was trained
to predict disease specific survival. The annotated gene expression data
and clinical information are available from the Gene Expression Omnibus
(identifier: GSE14814, \citet{edgar2002gene}).

\citet{zhu2010prognostic} present results that mainly focus on the
discrimination ability of their estimated signature. They do that by
demonstrating that the two risk subgroups predicted by their signature
(high risk and low risk) have separation in their survival curves and
that the hazard ratio for their signature is large and significant even
when adjusting for other risk factors. They do not directly address
calibration, that is, whether their signature accurately predicts
survival times.

We used a similar approach to preprocessing as did
\citet{zhu2010prognostic}, as we could not reproduce their workflow
exactly due to outdated software. Batch effects were removed using the
\texttt{ComBat} function in the \texttt{sva} \texttt{R} package
\citep{leeksva} and then the gene expression values were centered by
their means and scaled by their standard deviations. Our signature
development approach is similar but not identical to that in
\citet{zhu2010prognostic}. For purposes of illustrating the concepts we
used a simplyfied approach to signature development that retains the
main features of the original. The exact approach to signature
development does not have a major impact on the main conclusions of our
evaluation of the various approaches to signature performance
assessment.

@@

\section{Issues}\label{issues}

Recall that the main goal is to estimate
\(E_\mathcal{P}[\phi_\mathcal{F}(S)]\), the expected value of a given
statistic on future observations for \(f \in \mathcal{F}\). This can be
estimated with the in-sample empirical estimate:
\(\hat{E}[\phi_f(S)] = \frac{1}{n}\sum_{i=1}^n\phi_f(s_i)\) for a
particular \(f\). However, if \(S\) is used to estimate \(f\) then the
estimate will be biased due to overfitting, that is,
\(|E_\mathcal{P}[\phi_f(S)] - \hat{E}[\phi_f(S)]|\) will be large. This
bias results from the fact that \(\phi\) depends on \(f\), and thus the
statistic \(\phi\) is being adaptively defined based on the observed
data \(S\). Overfitting occurs when a model is fit to noise in the data.
This often occurs when fitting a model that is overly complex relative
to the amount of signal in the available data.

In many cases during signature development, the performance metric
\(\phi\) in question is a statistic that relates to calibration. While a
biomarker signature may accurately predict a clinical outcome, that does
not neccessarily imply that the signature is clinically useful. To
assess discrimination, a different statistic \(\phi\) may be used, such
as the area under the ROC curve. Measures of association such as the
odds ratio, hazard ratio, or difference in survival probabilities are
examples of performance metrics that are commonly used, although their
value for biomarker signatures has been debated
\citep{pepe2004limitations}. Evaluation of \(\phi\) is subject to bias
due to overfitting regardless of what \(\phi\) is used, a fact that is
commonly overlooked in the medical literature.

In \citet{zhu2010prognostic}, \(\phi\) was the hazard ratio comparing
the high risk and low risk groups as defined by the JBL.10 signature
\(f\). The risk groups were determined by the signature \(f\), which was
estimated using \(S\), the same data that was then used to produce the
hazard ratio estimate of 15. They go on to assess the performance of the
JBL.10 signature in a series of independent data sets, in which they
estimate hazard ratios of approximately 2. Here we reanalyze the dataset
and illustrate some remedies to avoid bias in estimating signature
performance.

\subsection{Avoiding Overfitting}\label{avoiding-overfitting}

A traditional method to avoid overfitting is the split sample approach.
First, randomly partition \(S\) into the training sample \(S_t\) and the
holdout sample \(S_h\) with sample sizes \(n_t\) and \(n_h\),
respectively. Then, \(S_h\) is hidden from the analyst while
\(\mathcal{F}\) is applied to \(S_t\) to estimate the signature function
\(f_t\). For fixed \(f_t\), \(\hat{E}[\phi_{f_t}(S_h)]\) is an unbiased
estimator of both \(E_{\mathcal{P}}[\phi_{f_t}(S)]\) and
\(E_\mathcal{P}[\phi_\mathcal{F}(S)]\). The specific form of \(f_t\)
that is fixed using the \(S_t\) partition can be reported as the
function for others to use, therefore the aforementioned estimator is
also an estimate of the error for that specific \(f_t\). The drawback of
the split-sample approach is that the performance of \(f_t\) is likely
to be inferior to the performance of a function \(f^*\) derived using
the same approach applied to the entire data set.
\citet{dobbin2011optimally} investigate how to optimally split a dataset
into training and holdout partitions.

Another approach to avoid overfitting is cross-validation, which is a
resampling based approach. For a fixed integer \(k\), which can be
between 1 and \(n\), we randomly select a partition of \(k\)
observations from \(S\), denoted \(S_k\). Then \(f_{-k}\) is estimated
and fixed by appling \(\mathcal{F}\) on \(S_{-k}\) which is the subset
of \(S\) that is disjoint from \(S_k\). Then, we get an estimate
\(\hat{E}[\phi_{f_{-k}}(S_k)]\) which is an unbiased estimate of
\(E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]\). This process is repeated
\(K\) times to yield \(K\) estimates. Each of these estimates is
unbiased, but noisy, because typically \(k\) is very small relative to
\(n\). Thus, we average over \(K\) to get a less noisy estimate. This
process is called ``leave \(k\) out'' or ``\(n/k\) fold''
cross-validation. Note that for each partition that is selected, we
obtain a new estimate of \(f\), therefore we are only estimating
\(E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]\) and not the performance for a
single specified signature. Typically, if a specific form for \(f\) is
desired, it would be estimated using the entire dataset \(S\) to yield
\(f^*\) as above.

A variation on the cross-validation approach is bootstrapping. In that
case, a sample \(S_b\) of size \(n\) is sampled \emph{with replacement}
from \(S\). Then \(f_{b}\) is estimated and fixed by applying
\(\mathcal{F}\) to \(S_{b}\). The performance metric \(\phi\) is
calculated on the subset of \(S\) that is disjoint from \(S_b\):
\(S_{-b}\) to yield an estimate of \(E[\phi_{f_b}(S_{-b})]\). This
process is repeated \(K\) times to yield \(K\) estimates. These \(K\)
estimates of the performance metrics are averaged to obtain the mean
over the bootstrap replicates. \citet{efron1997improvements} suggest a
variation, the 0.632 estimate: \[
\hat{E}^*[\phi_{\mathcal{F}}(S)] = .368 \hat{E}[\phi_{f}(S)] + 0.632 \hat{E}[\phi_{f_b}(S_{-b})],
\] where \(\hat{E}[\phi_{f}(S)]\) is the naive estimate of \(\phi_f\)
using the entire dataset.

Another variation on all of these methods is the concept of
pre-validation \citep{tibshirani2002pre}. With pre-validation, instead
of computing the statistic \(\phi\) for each of the held-out subsets
(\(S_{-b}\) for the bootstrap or \(S_{k}\) for cross-validation), the
fitted predictor \(\hat{f}(X_i)\) is estimated for \(X_i \in S_{-b}\)
where \(\hat{f}\) is estimated using \(S_{b}\). This process is repeated
to obtain a set of pre-validated predictor estimates \(\hat{f}\) which
are then used to calculate \(\phi\). For single-step holdout, this
process is equivalent to what is described above. For cross-validation
and the bootstrap, this process avoids the problem of having too few
cases to estimate the statistic \(\phi\) on the held-out dataset.

\subsection{Simulation Study}\label{simulation-study}

@@

To illustrate the different properties of these estimates and how they
deal with overfitting, we conduct a simulation study. Data were
generated with 1000 observations, each with a binary outcome \(Y\) with
prevalence 0.3, and 500 features sampled from the standard normal
distribution. This is the null case where no features are associated
with \(Y\). The signature development procedure entails a feature
selection step, in which each feature is regressed against \(Y\) in a
univariate logistic regression model. The 25 features with the smallest
p-values are selected for inclusion in a multivariable logistic
regression model which defines the final signature.

We compare each of the methods described above, split-sample holdout,
cross-validation, bootstrap, and pre-validation, along with several
commonly used but biased approaches. All of the approaches are described
in \ref{descript}. Two of the biased approaches use the full sample to
select the features, followed by fitting the multivariable model on the
holdout subset. This is referred to as ``parital holdout'' or ``partial
CV'' when using split-sample holdout or cross-validation as the
validation step, respectively. We also implemented the naive
resubstitution approach, wherein the model is trained and evaluated on
the same dataset, and the partial resubstitution approach wherein the
model is trained on a holdout set and then evaluated on the combined
complete dataset. Our main interest is in comparing the bias and
variance of the resulting estimates of
\(E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]\). In our simulation, we look
at two different performance metrics, the area under the ROC curve (AUC)
and the odds ratio for the outcome comparing the signature groups.

\begin{figure}[htbp]
\centering
\includegraphics{paper_files/figure-latex/cvsims-1.pdf}
\caption{Comparison of different approaches to estimating the Area Under
the ROC Curve (AUC) in the setting where a dataset is used to both
develop the signature and evaluate its performance. The violin plots
show mirrored density estimates for the AUC for 1000 replicates of the
numerical experiment. In each replicate, there are 1000 observations and
500 features. The true value of the AUC is 0.5. CV = Cross validation.
\label{fig1}}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{paper_files/figure-latex/cvsims2-1.pdf}
\caption{Comparison of different approaches to estimating the odds ratio
(OR) in the setting where a dataset is used to both develop the
signature and evaluate its performance. The violin plots show mirrored
density estimates for the log OR for 1000 replicates of the numerical
experiment. In each replicate, there are 1000 observations and 500
features. The true value of the OR is 1.0. CV = Cross validation.
\label{fig2}}
\end{figure}
